---
title: "Geographically Weighted Regression"
author: "Wen Fu"
date: "April 24, 2016"
output: 
  html_document: 
    keep_md: yes
---
Spatial non-stationarity: the same stimulus provokes a different response in different parts of the study region.

This lab tests and visualizes non-stationarity in a model aimed at explaining variations in cancer prevalence in central NY by regressing it on amount of exposure (exposure to toxic release site) and controlling for age (% of those over 65) and SES (% of home owners).

```{r}
suppressPackageStartupMessages(library(maptools))
suppressPackageStartupMessages(library(spdep))
suppressPackageStartupMessages(library(spgwr))
suppressPackageStartupMessages(library(rgdal))
```

Load the shapefile and plot it.
```{r}
NY_GWR <- readShapeSpatial("NY shapefile")
plot(NY_GWR)
```

See the variables associated with the shapefile.
```{r}
names(NY_GWR)
```

Here, "Z" is our dependent variable indicating standardized cancer rate. Our independent variables are "PEXPOSURE", "PCTAGE65P" and "PCTOWNHOME". View each of these variables spatially.
```{r}
spplot(NY_GWR, "Z")
spplot(NY_GWR, "PEXPOSURE")
spplot(NY_GWR, "PCTAGE65P")
spplot(NY_GWR, "PCTOWNHOME")
```

In order to examine for non-stationarity, need to identify the optimal bandwidth (bwG), which can be selected as the following.
```{r}
# Set weighting scheme as Gaussian (gweight) and ask for output to view (verbose = T)
bwG <- gwr.sel(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, data = NY_GWR, gweight = gwr.Gauss, verbose = TRUE) 
```

Now, use this bandwidth to produce the result - estimate the regression model with the newly created bandwidth, a Gaussian weighting scheme. View the results.
```{r}
gwrG <- gwr(Z ~ PEXPOSURE + PCTAGE65P + PCTOWNHOME, data = NY_GWR, bandwidth = bwG, gweight = gwr.Gauss)
gwrG
names(gwrG)
# View the names of the underlying spatial data frame (SDF)
names(gwrG$SDF) 
```

Visualize the geographical variation in the slope of exposure ($b$ of "PEXPOSURE") and model fit (local R squared) in the SDF.
```{r}
spplot(gwrG$SDF, "PEXPOSURE")
spplot(gwrG$SDF, "localR2")
```

If needed, write the results out to shapefile format for use in QGIS.
```{r, eval = F}
writeSpatialShape(gwrG$SDF, "GWR_Results")
```

